---
title: "Wisconsin Districts"
editor_options: 
  chunk_output_type: console
---

```{r}
#| message: false
#| include: false
library(fec16)

#library(fec12)
library(sf)        
library(ggspatial)
library(tidyverse)
library(ggplot2)
```


```{r}
#| message: false
#| include: false
district_elections <- results_house |>
  mutate(district = parse_number(district_id)) |>
  group_by(state, district) |>
  summarize(
    N = n(), 
    total_votes = sum(general_votes, na.rm = TRUE),
    d_votes = sum(ifelse(party == "DEM", general_votes, 0), na.rm = TRUE),
    r_votes = sum(ifelse(party == "REP", general_votes, 0), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    other_votes = total_votes - d_votes - r_votes,
    r_prop = r_votes / total_votes,  
    winner = ifelse(r_votes > d_votes, "Republican", "Democrat")
  )
wi_results <- district_elections |>
  filter(state == "WI")
wi_results |>                  
  select(-state)
```

```{r}
#| message: false
#| include: false

src <- "http://cdmaps.polisci.ucla.edu/shp/districts113.zip"
lcl_zip <- fs::path(tempdir(), "districts113.zip")
download.file(src, destfile = lcl_zip)
lcl_districts <- fs::path(tempdir(), "districts113")
unzip(lcl_zip, exdir = lcl_districts)
dsn_districts <- fs::path(lcl_districts, "districtShapes")
st_layers(dsn_districts)
districts <- st_read(dsn_districts, layer = "districts113") |>
  mutate(DISTRICT = parse_number(as.character(DISTRICT)))
wi_shp <- districts |>
  filter(STATENAME == "Wisconsin")
wi_merged <- wi_shp |>
  st_transform(4326) |>
  inner_join(wi_results, by = c("DISTRICT" = "district"))

# Color based on winning party
#   Note that geom_sf is part of ggplot2 package, while st_geometry is
#   part of sf package
wi <- ggplot(data = wi_merged, aes(fill = winner)) +
  annotation_map_tile(zoom = 6, type = "osm", progress = "none") + 
  geom_sf(alpha = 0.5) +
  scale_fill_manual("Winner", values = c("blue", "red")) + 
  geom_sf_label(aes(label = DISTRICT), fill = "white") + 
  theme_void()

wi
```

```{r}
#| message: false
#| echo: false
#| warning: false

wi +
  aes(fill = r_prop) + 
  scale_fill_distiller(
    "Proportion\nRepublican", 
    palette = "RdBu", 
    limits = c(0, 1)
  )
```

This map represents the proportion of people who voted for republican candidates in 2016 in each Wisconsin district. The only three districts that ended up winning democrat are 2, 3, and 4 which won by a lot since they are darker blues. There is evidence of gerrymandering in this map since all republican winning districts were intentionally made to barely win republican (which is seen in this map by the extremely slight tints of red for republican winning districts) so the state could come out winning republican overall. The gerrymandering was especially bad where districts 3 and 4 were so lopsided for democrats that they were such large blowouts that no republican candidates even ran since many democrats were intentionally lumped into these districts by the republicans that drew them so they could gain an advantage.
